% Created 2014-04-30 Wed 16:12
\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{fixltx2e}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{rotating}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{marvosym}
\usepackage{wasysym}
\usepackage{amssymb}
\usepackage{hyperref}
\tolerance=1000
\author{James Woodward Weis}
\date{\today}
\title{Compiled Biological Knowledge}
\hypersetup{
  pdfkeywords={},
  pdfsubject={},
  pdfcreator={Emacs 24.3.1 (Org mode 8.2.4)}}
\begin{document}

\maketitle
\tableofcontents

\section{Hallmarks of cancer}
\label{sec-1}

\subsection{Six original hallmarks of cancer}
\label{sec-1-1}
\begin{enumerate}
\item Sustain Proliferative Signaling
\item Evading Growth Suppressors
\item Resisting Cell Death
\item Enabling Replicative Immorality
\item Inducing Angiogenesis
\item Activating Invasion and Metastasis
\end{enumerate}
\subsection{Additional "enabling" and "emerging" hallmarks}
\label{sec-1-2}
\begin{enumerate}
\item Enabling: Genome instability and mutation
\item Enabling: Tumor-promoting inflammation
\item Emerging: Reprogramming energy metabolism
\item Emerging: Evading immune destruction
\end{enumerate}
\section{Intracellular circuitry of relevance in cancer}
\label{sec-2}
Circuit 'outputs' or results include:
\begin{itemize}
\item Gene Expression
\item Cell Proliferation
\end{itemize}
\subsection{Wnt signaling pathway (simplified)}
\label{sec-2-1}
\subsubsection{WNT + Frizzled -> Dishevelled}
\label{sec-2-1-1}
\subsubsection{Dishevelled -> GSK-3Beta}
\label{sec-2-1-2}
\subsubsection{GSK-3Beta -> APC}
\label{sec-2-1-3}
\subsubsection{APC -> Beta-Cetenin}
\label{sec-2-1-4}
\subsubsection{Other Cell + E-Cadherin -> Beta-Catenin}
\label{sec-2-1-5}
\subsubsection{Beta-Catenin + TCF -> Changes in Gene Expression}
\label{sec-2-1-6}
**

\section{References}
\label{sec-3}
\begin{itemize}
\item Hanahan, D., and Weinberg, R.A. (2000). The Hallmarks of Cancer. Cell 100, 57-70.
\item Hanahan, D., and Weinberg, R.A. (2011). The Hallmarks of Cancer: The Next Generation. Cell 144, 646-674.
\item Huelsken, J., and Juergen, B. (2000). The Wnt signalling pathway. J. Cell Sci. 113, 3545.
\end{itemize}



\begin{itemize}
\item Names. (date) title. journal issue, page-page.
\end{itemize}
% Emacs 24.3.1 (Org mode 8.2.4)
\end{document}